---
title: "Lab 10"
author: "Gretel Warmuth (PID: A17595945)"
format: pdf
---

## Importing candy data

Uploading candy data from 538 into this R project:

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
candy
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

```{r}
rownames(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

```{r}
sum(candy$chocolate)
```


## What is your favorate candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Milky Way", ]$winpercent
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```


```{r}
library(dplyr)

candy |>
  filter(rownames(candy) == "Haribo Happy Cola") |> 
  select(winpercent)
```



> Q. Find fruity candy with a win percent above 50%

Using the filter feature: 

```{r}
candy |> 
  filter(winpercent > 50) |>
  filter(fruity == 1)
```

```{r}
rownames(candy |> 
  filter(winpercent > 50) |>
  filter(fruity == 1))
```

Another way:

```{r}
top.candy <- candy[candy$winpercent > 50, ]
top.candy[top.candy$fruity == 1, ]
```

```{r}
rownames(top.candy[top.candy$fruity == 1, ])
```

Usng skim to read the data:

```{r}
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The `winpercent` column is measured differently than the rest of the other variables. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

The represent a "true" (1) or "false" (0) value.

> Q8. Plot a histogram of winpercent values

Using "base" R `hist()` function:

```{r}
hist(candy$winpercent, breaks = 20)
```

Using ggplot:

```{r}
library(ggplot2)
ggplot(candy, aes(x = winpercent, col = "orange")) +
geom_histogram(binwidth = 10) +
  theme_light()
```


> Q9. Is the distribution of winpercent values symmetrical?

No

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
fruit.candy <- candy |> 
  filter(fruity == 1)

summary(fruit.candy$winpercent)
```

```{r}
chocolate.candy <- candy |>
  filter(chocolate == 1)

summary(chocolate.candy$winpercent)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(chocolate.candy$winpercent, fruit.candy$winpercent)
```

Small p-value means there is a statistically significant difference between fruity and chocolate candy. 

## Overall Candy Rankings

Learning to sort and order:

```{r}
play <- c("d", "a", "c")
sort(play)
order(play)

play[order(play)]
```


```{r}
sort(c(5,2,10), decreasing = T)
```

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], 5)
```

```{r}
rownames(head(candy[order(candy$winpercent),], 5))
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing = T),], 5)
```

```{r}
rownames(head(candy[order(candy$winpercent, decreasing = T),], 5))
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy, aes(x = winpercent, y = rownames(candy))) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy, aes(x = winpercent, y = reorder(rownames(candy), winpercent), fill = chocolate)) +
  geom_col()
```

There needs to be a better color scheme to show the chocolate, fruity, caramel, etc categories. Using custom vector:

```{r}
# Placr holder color vector

mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] <- "pink"

rownames(candy) == "Milky Way"

mycols[rownames(candy) == "Milky Way"] <- "blue"
mycols
```

```{r}
ggplot(candy, aes(x = winpercent, y = reorder(rownames(candy), winpercent))) +
  geom_col(fill = mycols)
```

> Q17. What is the worst ranked chocolate candy?

Sixlet

> Q18. What is the best ranked fruity candy?

Starburst

## Taking a look at pricepercent

Plotting winpercent vs pricepercent to see the best candy to buy:

```{r}
mycols[as.logical(candy$fruity)] <- "red"
```


```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 30)

```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Lip, Ring Pop, Nestle Smarties, Mr. Goodbar, and Milky Way. Nik L Lip is the least popular.

## Exploring the correlation structure

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij, diag = F, type = "upper")
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate/fruity and plurbius/bar

> Q23. Similarly, what two variables are most positively correlated?

winpercent/chocolate and bar/chocolate

## Principal Component Analysis

```{r}
library(stats)
pca <- prcomp(candy, scale = T)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2], col = mycols, pch = 16)
```

Showing how the original variables contribute to the new PCs. PC1:

```{r}
loadings <- as.data.frame(pca$rotation)

ggplot(loadings, aes(x = PC1, y = reorder(rownames(loadings), PC1), fill = PC1)) +
  geom_col()
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data, aes(x = PC1, y = PC2, size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data))) +
  geom_point(col = mycols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 40)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#library(plotly)
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and plurbius

