---
title: "Lab 8"
author: "Gretel Warmuth (PID: A17595945)"
format: gfm
---

# PCA Mini Project

It is important to consider scalling your data before analysis such as PCA.

For example:

```{r}
head(mtcars)
```

```{r}
colMeans(mtcars)
```

```{r}
apply(mtcars, 2, sd)
```

```{r}
x <- scale(mtcars)
head(x)
```

```{r}
round(colMeans(x), 2)
```

## Preparing the Data

Downloading, preparing, and reading our data for breast cancer: 

```{r}
fna.data <- "https://bioboot.github.io/bimm143_S20/class-material/WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
head(wisc.df)
```

We need to omit the first column `diagnosis` so that PCA doesn't read this column. We will use the column as a comparison to our own analysis:

```{r}
# We can use -1 to remove the first column

wisc.data <- wisc.df[,-1]
head(wisc.data)
```

For future use, we will need a factor that includes the `diagnosis` column:

```{r}
# We can use the table() function to tell us how many malignant and benign cells there are in the data

diagnosis_table <- wisc.df[, 1]
table(diagnosis_table)
```

```{r}
diagnosis <- as.factor(wisc.df[, 1])
head(diagnosis)
```

## Exploratory Data Analysis

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```


> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df[, 1])
```


> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("mean", names(wisc.df), value = T, fixed = F))
```


## Principal Component Analysis

Checking means and standard deviations:

```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

Scaling the Data:

```{r}
wisc.pr <- prcomp( wisc.data, scale = T )
summary(wisc.pr)
```

What's in the PCA result object?

```{r}
attributes(wisc.pr)
```


```{r}
head(wisc.pr$x)
```


Main "PC Score Plot"/"PC1 vs PC2 Plot":

```{r}
# plot(wisc.pr$x)
plot(wisc.pr$x[, 1], wisc.pr$x[, 2], col = as.factor(diagnosis))
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

```{r}
summary(wisc.pr)
```


> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?


## Interpreting PCA results

Creating a biplot:

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

Creating a scatter plot:

```{r}
plot(wisc.pr$x, col = diagnosis, xlab= "PC1", ylab = "PC2")
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(wisc.pr$x[, 1:3 ], col = diagnosis, 
     xlab = "PC1", ylab = "PC3")
```

Creating a ggplot of the data:

```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

## Variance explained

Calculating the variance: 

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
# pve <- pr.var / 

# Plot variance explained for each principal component
# plot(pve, xlab = "Principal Component", 
     # ylab = "Proportion of Variance Explained", 
     # ylim = c(0, 1), type = "o")
```

## Communicating PCA results

>Q. Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?



>Q. Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

## 3. Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
head(data.scaled)
```

```{r}
data.dist <- dist(data.scaled)
head(data.dist)
```

```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
wisc.hclust
```

## Results of hierarchical clustering

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(wisc.hclust, col="red", lty=2)
```

```{r}
# distance matrix from PCA result
d <- dist(wisc.pr$x[, 1:3])
hc <- hclust(d, method = "ward.D2")
plot(hc)
```

```{r}
grps <- cutree(hc, k = 2)
```

```{r}
plot(wisc.pr$x, col = grps)
```

Compare clustering result to expert diagnosis

```{r}
table(diagnosis, grps)
```




